\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-04-27 main.tex
draft donation section

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.65\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_selection_strenght.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_selection_strenght.pdf}
  \end{subfigure}%
  \begin{subfigure}{.35\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different benefit values.} 
Here, we vary the benefit of cooperation $b$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. Unless explicitly
varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$,
$\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$
time steps for each parameter
combination.}\label{fig:cooperation_rate_over_benefit}
\end{figure}

@@

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_beta.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_beta.pdf}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different parameter values.} 
While the previous figures depict the evolutionary outcome for fixed parameter
values, here we vary the benefit of cooperation $b$ and the strength of
selection $\beta$. In all cases, stochastic payoff evaluation tends to reduce
the evolving cooperation rates. Unless explicitly varied, the parameters of the
simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$,
$\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$ time steps for
each parameter combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

@@

In evolution context we consider a population of \(N\) players,
where \(N\) is even, and where mutations are sufficiently rare. At any point
in time the there are at most two different strategies are present in
the population. Suppose there are \(N - k\) players who use the strategy \(s_{1}=(y_{1},
p_{1}, q_{1})\), whereas \(k\) players use the strategy \(s_{2}=(y_{2}, p_{2},
q_{2})\). We refer to these two player types as ‘residents’ and ‘mutants’,
respectively.

Each step of the evolutionary process consists of two stages, a game stage and
an updating stage.

\begin{enumerate}
    \item In the game stage, each player is randomly matched with some
    other player in the population to interact in one instance of the IPD.
    \item In the updating stage, two players are randomly drawn from
    the population, a `learner' and a `exemplar'. Given that the learner's payoff
    in the last round is $u_L\!\in\! \mathcal{U}$ and that the exemplar's last
    round's payoff $u_E\!\in\! \mathcal{U}$, we assume the learner adopts the role
    model's strategy with probability

    \begin{equation} \label{Eq:rho}
    \rho(u_L, u_E) = \frac{1}{1\!+\!\exp\big[ \!-\!\beta (u_E\!-\!u_L) \big]}. 
    \end{equation}

    where $\beta\!\ge\!0$ corresponds to the strength of selection.
\end{enumerate}

We iterate this basic evolutionary step until either the mutant strategy goes
extinct, or until it fixes in the population (in which case the mutant strategy
becomes the new resident strategy). After either outcome, we introduce a new
mutant strategy $s'_2\!=\!(y_2',p_2',q_2')$ (uniformly chosen from all reactive
strategies at random), and we set the number of mutants to $k\!=\!1$. This
process of mutation and fixation/extinction is then iterated many times.

\section{Expected Payoffs}

We compare this process for what we defined as {\bf stochastic payoff
evaluation} with the analogous process where players update their strategies
with respect to their {\bf expected} payoffs,

\begin{equation} \label{eq:ExpPay}
\begin{array}{lcrcr}
\pi_1 &= &\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \langle\mathbf{v}(s_1,s_1),\mathbf{U}\rangle	& + &\displaystyle\frac{k}{N-1}\cdot \langle\mathbf{v}(s_1,s_2),\mathbf{U}\rangle,\\[0.5cm]
\pi_2 &= &\displaystyle\frac{N-k}{N-1}\cdot \langle\mathbf{v}(s_2,s_1),\mathbf{U}\rangle & + &\displaystyle\frac{k-1}{N-1}\cdot \langle\mathbf{v}(s_2,s_2),\mathbf{U}\rangle.\\
\end{array}
\end{equation}

@@

\section{Stochastic Payoffs}

We define {\bf stochastic payoff} as the average payoff $u\!\in\! \mathcal{U}$ a player
receives in the last \(n\) rounds of the game given that they interact with
\(m\) players.

\subsubsection*{Case \(n=m=1\).}

Initially, consider the situation where \(n=m=1\). The player's stochastic
payoff is what they receive in the last round against a single opponent. There
only four possible outcomes for the last round, those are \(CC, CD, DC, DD\).
Consider two players with reactive strategies $S_1\!=\!(y_1, p_1, q_1$) and
$S_2\!=\!(y_2,p_2,q_2)$ who interact in a repeated prisoner's dilemma with
continuation probability $\delta$, the probability that are in each of the
four possible states in the last round is given by:

@@

Equation~\ref{Eq:LastRound} is the probability vector that players \(S_1, S_2\)
are in each of the possible states of the last round. Given the population \(N\)
with \(k\) mutants in the last round there only possible combinations of interactions
are:

@@

We compare this process for stochastic payoff evaluation with the analogous process where players update their strategies with respect to their {\it expected} payoffs,
\begin{equation} \label{Eq:ExpPay}
\begin{array}{lcrcr}
\displaystyle \pi_1	&=	&\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \langle\mathbf{v}(S_1,S_1),\mathbf{u}\rangle	&+	&\displaystyle\frac{k}{N-1}\cdot \langle\mathbf{v}(S_1,S_2),\mathbf{u}\rangle,\\[0.5cm]
\displaystyle \pi_2	&=	&\displaystyle\frac{N-k}{N-1}\cdot \langle\mathbf{v}(S_2,S_1),\mathbf{u}\rangle&+	&\displaystyle\frac{k-1}{N-1}\cdot \langle\mathbf{v}(S_2,S_2),\mathbf{u}\rangle.\\
\end{array}
\end{equation}
In the limit of no discounting, $\delta\!\rightarrow\! 1$, this process based on expected payoffs has been considered in~\cite{imhof2010stochastic}.

\subsection{Fixation probabilities under stochastic payoff evaluation}
Given that $N\!-\!k$ players use the resident strategy $S_1\!=\!(y_1,p_1,q_1)$ and that the remaining $k$ players use the mutant strategy $S_2\!=\!(y_2,p_2,q_2)$, the probability that the number of mutants increases by one in one step of the evolutionary process can be written as
\begin{equation}
\lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_1,u_2\in\mathcal{U}} x(u_1,u_2)\cdot \rho(u_1,u_2).
\end{equation}
In this expression, $(N\!-\!k)/N$ is the probability that the randomly chosen learner is a resident, and $k/N$ is the probability that the role model is a mutant. The sum corresponds to the total probability that the learner adopts the role model's strategy over all possible payoffs $u_1$ and $u_2$ that the two player may have received in their respective last rounds. We use $x(u_1,u_2)$ to denote the probability that the randomly chosen resident obtained a payoff of $u_1$ in the last round of his respective game, and that the mutant obtained a payoff of $u_2$. Given that the payoffs are $u_1$ and $u_2$, the imitation probability is then given by $\rho(u_1,u_2)$, as specified by Eq.~(\ref{Eq:rho}). The probability that the respective payoffs of the players are given by $u_1$ and $u_2$ can be calculated as
\begin{equation}
\setlength{\arraycolsep}{1pt}
\begin{array}{llrl}
x(u_1,u_2)	 &=&\displaystyle \frac{1}{N\!-\!1}\cdot  &v_{u_1}(S_1,S_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F}\\[0.5cm]
&+	
&\displaystyle \left(1\!-\!\frac{1}{N\!-\!1}\right)  
&\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_2) + 
 \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_1)\right.\\[0.5cm]
&&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_2) + 
 \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_1)\right].
\end{array}
\end{equation}
The first term on the right side corresponds to the case that the learner and the role model happened to be matched during the game stage, which happens with probability $1/(N\!-\!1)$. In that case, we note that only those payoff pairs can occur that are feasible in a direct interaction, $(u_1,u_2)\in \mathcal{U}^2_F:=\big\{ (R,R), (S,T), (T,S), (P,P) \big\}$, as represented by the respective indicator function. Otherwise, if the learner and the role model did not interact directly, we need to distinguish four different cases, depending on whether the learner was matched with a resident or a mutant, and depending on whether the role model was matched with a resident or a mutant.

>>>>

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.65\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_selection_strenght.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_selection_strenght.pdf}
  \end{subfigure}%
  \begin{subfigure}{.35\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different benefit values.} 
Here, we vary the benefit of defection $b$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. Unless explicitly
varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$,
$\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$
time steps for each parameter
combination.}\label{fig:cooperation_rate_over_benefit}
\end{figure}

@@

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_beta.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_beta.pdf}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different selection strength values.} 
Here, we vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. Unless explicitly
varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$,
$\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$
time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

@@

In the game stage, each player is randomly matched with some other player in
the population to interact in a number of instances of the game
\(\begin{pmatrix} R & S  \\ T & P \end{pmatrix}\). The number of interactions is
not fixed, on the contrast we consider that another interaction has a
probability $\delta$ to continue following each turn. We assume herein that
individuals at most make use of simple \textit{reactive strategies} make
decisions in each round. Reactive strategy are a set of memory-one strategies
that only take into account the previous action of the opponent. Reactive
strategies can be written explicitly as a vector \(\in \R_{3}\). More
specifically a reactive strategy \(s\) is given by \(s=(y, p, q)\) where \(y\)
is the probability that the strategy opens with a cooperation and \(p, q\) are
the probabilities that the strategy cooperates given that the opponent
cooperated and defected equivalently.

In the updating stage, two players are randomly drawn from the population, a
`learner' and a `role model'. Given that the learner's payoff $u_L\!\in\!
\mathcal{U}$ and that the role model's payoff $u_{RL}\!\in\! \mathcal{U}$, we
assume the learner adopts the role model's strategy with probability

\begin{equation} \label{Eq:rho}
\rho(u_L, u_{RL}) = \frac{1}{1\!+\!\exp\big[ \!-\!\beta (u_E\!-\!u_{RL}) \big]}.
\end{equation}

where $\beta\!\ge\!0$ corresponds  to relative influence of the payoffs on the
adopt the strategy of the other is controlled by an external parameter, the so
called intensity of selection.

We iterate this basic evolutionary step until either the mutant strategy goes
extinct, or until it fixes in the population (in which case the mutant strategy
becomes the new resident strategy). After either outcome, we introduce a new
mutant strategy (uniformly chosen from all reactive
strategies at random), and we set the number of mutants to $k\!=\!1$. This
process of mutation and fixation/extinction is then iterated many times.xw The
fixation probability of the mutant strategy then takes the standard form
\cite{nowak2004emergence},

\begin{equation}
\varphi = \frac{1}{1+\sum_{i=1}^{N-1}\prod_k^i \frac{\lambda^-_k}{\lambda^+_k}}.
\end{equation}

where \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of
mutants decreases and increases respectively.

We compare this process for stochastic payoff evaluation with the analogous
process where players update their strategies with respect to their {\it
expected} payoffs.

\subsection{Expected Payoffs}

Consider two players with reactive strategies $s_1\!=\!(y_1, p_1, q_1$) and
$s_2\!=\!(y_2,p_2,q_2)$ who interact in a repeated prisoner's dilemma with
continuation probability $\delta$, the probability that are in each of the
four possible states in the last round is given by the stationary distribution
\(\langle\mathbf{v}\) of the transition matrix \(M\).

\begin{equation}\label{eq:transition_matrix}
  \input{static/matrix.tex}.
\end{equation}

The long run steady state probability vector \(\langle\mathbf{v}\) is the solution
to \(\langle\mathbf{v}M=\langle\mathbf{v}\), can be combined with the payoff
\(u\) to give the payoffs for each player. More specifically,

\[\langle\mathbf{v}(s_1,s_2),\mathbf{u}\rangle\]

is the payoff that \(s_1\) achieves when interacting with \(s_2\). In
the evolutionary process given \(k\) individuals with mutant strategy \(s_M\)
and \(N - k\) individuals with the resident strategy \(s_R\) the expected
payoffs are,

\begin{equation} \label{Eq:ExpPay}
  \begin{array}{lcrcr}
  \displaystyle \pi_R	&=	&\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \langle\mathbf{v}(s_R,s_R),\mathbf{u}\rangle	&+	&\displaystyle\frac{k}{N-1}\cdot \langle\mathbf{v}(s_R,s_M),\mathbf{u}\rangle,\\[0.5cm]
  \displaystyle \pi_M	&=	&\displaystyle\frac{N-k}{N-1}\cdot \langle\mathbf{v}(s_M,s_R),\mathbf{u}\rangle&+	&\displaystyle\frac{k-1}{N-1}\cdot \langle\mathbf{v}(s_M,s_M),\mathbf{u}\rangle.\\
  \end{array}
\end{equation}

@@

\subsection{Stochastic Payoffs}

Initially, we consider the case that an individual's fitness is based on their
last interaction with one other individual. There
only four possible outcomes for the last round, those are \(CC, CD, DC, DD\).
Consider two players with reactive strategies $s_1\!=\!(y_1, p_1, q_1$) and
$s_2\!=\!(y_2,p_2,q_2)$ who interact in a repeated prisoner's dilemma with
continuation probability $\delta$, the probability that are in each of the
four possible states in the last round is given by:

@@

Equation~\ref{Eq:LastRound} can be generalised to all cases where we consider
the last \()

Equation~\ref{Eq:LastRound} is the probability vector that players \(s_1, s_2\)
are in each of the possible states of the last round. Given the population \(N\)
with \(k\) mutants in the last round there only possible combinations of interactions
are:

@@

\subsection{Fixation probabilities under stochastic payoff evaluation}
Given that $N\!-\!k$ players use the resident strategy $S_1\!=\!(y_1,p_1,q_1)$ and that the remaining $k$ players use the mutant strategy $S_2\!=\!(y_2,p_2,q_2)$, the probability that the number of mutants increases by one in one step of the evolutionary process can be written as
\begin{equation}
\lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_1,u_2\in\mathcal{U}} x(u_1,u_2)\cdot \rho(u_1,u_2).
\end{equation}
In this expression, $(N\!-\!k)/N$ is the probability that the randomly chosen learner is a resident, and $k/N$ is the probability that the role model is a mutant. The sum corresponds to the total probability that the learner adopts the role model's strategy over all possible payoffs $u_1$ and $u_2$ that the two player may have received in their respective last rounds. We use $x(u_1,u_2)$ to denote the probability that the randomly chosen resident obtained a payoff of $u_1$ in the last round of his respective game, and that the mutant obtained a payoff of $u_2$. Given that the payoffs are $u_1$ and $u_2$, the imitation probability is then given by $\rho(u_1,u_2)$, as specified by Eq.~(\ref{Eq:rho}). The probability that the respective payoffs of the players are given by $u_1$ and $u_2$ can be calculated as
\begin{equation}
\setlength{\arraycolsep}{1pt}
\begin{array}{llrl}
x(u_1,u_2)	 &=&\displaystyle \frac{1}{N\!-\!1}\cdot  &v_{u_1}(S_1,S_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F}\\[0.5cm]
&+	
&\displaystyle \left(1\!-\!\frac{1}{N\!-\!1}\right)  
&\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_2) + 
 \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_1)\right.\\[0.5cm]
&&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_2) + 
 \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_1)\right].
\end{array}
\end{equation}

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the game stage, which happens with
probability $1/(N\!-\!1)$. In that case, we note that only those payoff pairs
can occur that are feasible in a direct interaction, $(u_1,u_2)\in
\mathcal{U}^2_F:=\big\{ (R,R), (S,T), (T,S), (P,P) \big\}$, as represented by
the respective indicator function. Otherwise, if the learner and the role model
did not interact directly, we need to distinguish four different cases,
depending on whether the learner was matched with a resident or a mutant, and
depending on whether the role model was matched with a resident or a mutant.
